Robust Digital Envelope Estimation Via Geometric Properties of an Arbitrary Real Signal

### 2020-09-04 Papers/02_Pulses_Envelope/PulsesEnvelope.tex
some progress with the paper

\begin{abstract} % 250 words
% background  
Despite being an elusive concept, the temporal amplitude envelope of a signal is essential for its complete characterization, being the primary information carrying medium in spoken voice and telecommunications, for example.
intuitively, the temporal envelope can be understood as a smooth function, that modulates the signal, being responsible for its outer shape.
% motivation
Envelope detection techniques have applications in areas like health, sound classification and synthesis, seismology and speech recognition. Despite of that, a general approach to envelope detection of signals with rich spectral content seems to be lacking, and most methods involve manual intervention, like filter design or smoothing, based on prior knowledge about the wave under investigation.
% objective
In this paper we propose a framework that uses intrinsic characteristics of a signal to estimate its envelope, completely eliminating the necessity of parameter tuning.
% methods
The approach here described draws inspiration from general, computational and differential geometry to isolate the frontiers and thus estimate the temporal envelope of an arbitrary signal; to that end The concepts of alpha-shapes, concave hulls, and discrete curvature are explored. We also define a pulse in the context of a arbitrary digital signal as a means to reduce dimensionality and the complexity of the proposed algorithm.
% results
We find that the algorithm accurate in the identification of the frontiers of a range of digital sound signal with very diverse characteristics, while providing the dominant frequency of the wave as a by-product. The algorithm also compares favourably with a common detection technique based on the Hilbert Transform.
While the results are satisfactory per se, we define several geometric entities in this work, including a new measure of discrete curvature; we hope to consolidate theory in the area, while providing a solid basis for the development of new techniques, specially in sound synthesis.
\end{abstract}

@@

\section{Introduction} % 500 words
% Opener sentence
Despite being ubiquitous in both analog and digital signal processing, the literature about envelope detection is very fragmented \parencite{2017LyonsDigital}. Besides, most envelope detection techniques are designed to account for very specific settings, like pure sinusoids with moderate noise contents, in line with the most common usages of those algorithms with artificial signals in the context of analog telecommunications; that limitation arises in part by the lack of a mathematical definition for a temporal envelope \parencite{2013Mengempirical}.

% Literature review | Brief Context of Prior Research
In many natural signals, however, the temporal amplitude envelope of a signal plays a prominent role in the characteristics exhibited: According to \textcite{2017QiRelative}, for example, the envelope is at least as important as the fine structure of the soundwave in the context of the intelligibility of mandarin tones. That is also the case for English \parencite{1995ShannonSpeech}, where even envelopes modulating mostly noise were still capable of conveying meaning. The envelope helps to convey emotion and identity to the human voice \parencite{2018ZhuContributions}, and envelope preserving characteristics of concert halls are associated with their pleasantness \parencite{2011LokkiEngaging}.

@@

@@

% State the Significance of Your Question
While the robustness of the proposed approach allows it to be used as a plug in replacement for many methods encountered in the literature, we feel that it would be particularly useful for sound synthesis. 
The envelope is shown to add complexity to the spectral representation of a wave \parencite{2019TarjanoNeuro}, and a accurate description of the envelope would be useful for a cleaner spectral analysis. Moreover, the algorithm naturally divides a signal in pseudo-cycles, that could serve pragmatic building blocks for the reconstruction of the fine structure of the wave.

% State Your Claim | Objective | Hypothesis
In the context of sound synthesis, one would be able to apply specific methods for the recreation of the envelope, more in line with its smooth, slow varying and non periodic nature and a different approach to the periodic and relatively fast changes of the temporal fine structure.
% TODO brief outlook on the structure of the paper

@@

% Essential background information
We start by defining a pulse as a series of consecutive samples in a discrete wave with the same sign. More formally, let $ W[i] \in \mathbb{R} \forall i \in \mathbb{N}_0, i < n \in \mathbb{N}_0 $ be a real, discrete and finite signal indexed, without loss of generality, over a subset of the natural numbers. This definition relates closely to the concept of an array or vector in most programming languages, and is used in the interest of simplicity.

@@

In other words, each pulse is a unique subset of $ W $ with its members being the maximum sequence of samples in $ W $ with the same signal and $ P[0] \cup P[0] \cup \cdots \cup P[q-1] = W $. It is convenient to attribute to each pulse a pair of coordinates: those will be the coordinates of the sample with the maximum amplitude in each pulse.

Naturally, each pulse can be classified as positive or negative, based on the signal of its samples, which must be the same.
For a continuous function, then, a pulse would correspond to the pieces between it's roots, and would be equal to half the cycle of a sinusoid. Thus, in our analogy, two consecutive pulses, naturally with opposing signs, have the potential do delimitate a pseudo-cycle, as illustrated in figure \ref{fig:Pulses}. For that to be true, however, the pulses need to make part of the frontiers of the signal, as will be defined later.

@@

To do so we will define the positive (upper) frontier and negative (lower) frontier as the set of points $ (i, |W[i]|) $ that are, necessarily but not sufficiently, the point of maximum amplitude of their respective pulses. The set of all these points such that $ W[i] > 0 $ will be called the positive, piecewise linear frontier and will be denoted by $ F_+ $; similarly, $ F_- $ will denote the negative frontier of $ W $, defined by the subset of points $ (i, W[i]) $ where $ W[i] < 0 $, as is shown in figure \ref{fig:Frontiers}. To identify such points, we need to address the different units in which the amplitude and horizontal distance of our wave are expressed.

@@

To make abscissa and ordinate units consistent for the computation of the frontier, and proceed with a geometric treatment of the subject, we consider the set of positive and negative pulses separately. For each set, we multiply all pulses by the average length of the pulses in the set, after having divided by the average of the pulses' amplitudes, putting both axes in the same frame related unit $i$, related to time by the equation $i = t \ \mathit{fps}$.

@@

Since an edge defines only one direction, there remains the need for a reference. As we are interested in the relative curvature, it is natural to use the average direction of all vectors, $ \vec{\overline{v}}^\pm = \frac{1}{m-1} \sum_{l=0}^{m-1} \vec{v}_l^\pm $, so as to measure the relative change of direction of each edge. This procedure, as well as the circles generated by it, are illustrated in figure \ref{fig:Curvatures}.

The angle $ \theta $ between the average direction and each of the vectors can be obtained via the relation of the slope of the respective lines, as in the expression $ \theta = \arctan( (m - \overline{m}) / (1 + m * \overline{m}) ) $. The radii of the equivalent circles are thus given by $ r_l = x_l / \sin (\theta_l) $ with the curvature given by $\kappa_l =1/r_l=\sin (\theta_l) / x_l$.

@@

% Therefore, the definition of the curvature becomes $ \kappa(x) = \frac{2 a}{\big( (b + 2 a x)^2 + 1 \big)^\frac{3}{2}} $ with the average curvature between points $ x_0 $ and $ x_1 $ given by equation \ref{eq:1}.

% \begin{equation} \label{eq:1}
%   \overline{\kappa}(x) = 
%   \frac{
%       \left( \dfrac{2 a x_1 + b}{\sqrt{\big(2 a x_1 + b \big)^2+1}} - \dfrac{2 a x_0 + b}{\sqrt{\big(2 a x_0 + b \big)^2+1}} \right) 
%   }{
%     (x_1 - x_0)
%   }
% \end{equation}

% \subsection{Theory}

\section{Results}
% TODO
% Revisit purpose and methods
% Overview of results
Figure \ref{fig:FullFrontiers} illustrates the frontiers obtained to six diverse discrete sound waves, as well as a detail view of a shorter segment for each wave. All waves are records of physical sounds, chosen to represent the applicability of the algorithm in real world scenarios.

@@

\begin{figure}[ht!]
  \centering
    \includegraphics[width=0.8\linewidth]{07FullFrontiers.pdf}
  \caption{Positive and negative frontiers of 6 digital waves. For each wave, the region highlighted in black is shown in detail. For each wave the horizontal axis is the sample number $i$, while the vertical axis is the normalized amplitude. All 6 waves were recorded at 44100 $ \mathit{fps} $.}
  \label{fig:FullFrontiers}
\end{figure}

@@

\subsection{Fundamental Frequency as a Welcome Side-effect}
% Comparison with results in other works
% Comparison with predictions
% Problems with results
% Possible implications

\section{Discussion + Conclusion}

\section{References}
\printbibliography[heading=none]

>>>>

\begin{abstract} % 250 words
% background  
Despite being an elusive concept, the temporal amplitude envelope of a signal is essential for its complete characterization, being the primary information carrying medium in spoken voice and telecommunications, for example.
intuitively, the temporal envelope can be understood as a smooth function, that modulates the signal, being responsible for its outer shape.
% motivation
Envelope detection techniques have applications in areas like health, sound classification and synthesis, seismology and speech recognition. Despite of that, a general approach to envelope detection of signals with rich spectral content seems to be lacking, and most methods involve manual intervention, like filter design or smoothing, based on prior knowledge about the wave under investigation.
% objective
In this paper we propose a framework that uses intrinsic characteristics of a signal to estimate its envelope, completely eliminating the necessity of parameter tuning.
% methods
The approach here described draws inspiration from general, computational and differential geometry to isolate the frontiers and thus estimate the temporal envelope of an arbitrary signal; to that end The concepts of alpha-shapes, concave hulls, and discrete curvature are explored. We also define a pulse in the context of a arbitrary digital signal as a means to reduce dimensionality and the complexity of the proposed algorithm.
% results
We find the algorithm accurate in the identification of the frontiers of a wide range of digital sound signal with very diverse characteristics, while providing the dominant frequency of the wave as a by-product. The algorithm also compares favourably with a common detection technique based on the Hilbert Transform.
While the results are satisfactory per se, we define several geometric entities in this work, including a new measure of discrete curvature; hoping to consolidate theory in the area, while providing a solid basis for the development of new techniques, specially in sound synthesis and classification.
\end{abstract}

@@

\section{Introduction} % 500 words
% Opener sentence
Despite being ubiquitous in both analog and digital signal processing \parencite{2011CaetanoImproved}, the literature about envelope detection is very fragmented \parencite{2017LyonsDigital}. Besides, most envelope detection techniques are designed to account for very specific settings, like pure sinusoids with moderate noise contents, in line with the most common usages of those algorithms with artificial signals in the context of analog telecommunications; that limitation arises in part by the lack of a mathematical definition for a temporal envelope \parencite{2013Mengempirical}.

% Literature review | Brief Context of Prior Research
In many natural signals, however, the temporal amplitude envelope of a signal plays a prominent role in the characteristics exhibited: According to \textcite{2017QiRelative}, for example, the envelope is at least as important as the fine structure of the soundwave in the context of the intelligibility of mandarin tones. That is also the case for English \parencite{1995ShannonSpeech}, where even envelopes modulating mostly noise were still capable of conveying meaning.

The envelope helps to convey emotion and identity to the human voice \parencite{2018ZhuContributions}, and envelope preserving characteristics of concert halls are associated with their pleasantness \parencite{2011LokkiEngaging}.

@@

Interesting steps in the direction of translating geometric algorithms to the context of envelope detection were made by \textcite{2015YangSkeleton} with an approach based on the construction of a skeleton underlying the wave, and with the direct translation of computer vision methods to the task \parencite{2015YangRepresenting}.

@@

% State the Significance of Your Question
While the robustness of the proposed approach allows it to be used as a plug in replacement for many methods encountered in the literature, we feel that it would be particularly useful for sound synthesis; the envelope is shown to add complexity to the spectral representation of a wave \parencite{2019TarjanoNeuro}, and a accurate description of the envelope would be useful for a cleaner spectral analysis. Moreover, the algorithm naturally divides a signal in pseudo-cycles, that could serve pragmatic building blocks for the reconstruction of the fine structure of the wave.

% State Your Claim | Objective | Hypothesis
In the context of sound synthesis, for example, one would be able to apply specific methods for the recreation of the envelope, more in line with its smooth, slow varying and non periodic nature and a different approach to the periodic and relatively fast changes of the temporal fine structure.

% brief outlook on the structure of the paper
After explaining and defining the various entities used later, of which the concept of frontiers is arguably the most important, in the methodology section, we dedicate a section to present a new discrete curvature estimation approach.
We proceed to illustrate the results of the application of the algorithm in a set of 6 diverse sound waves, ranging from voice to unpitched instruments.

We then discuss the future implications of the work, emphasizing the focus on the construction of new interpretations of the relation between time domain and frequency domain sound descriptions, and our vision of the impact of this methodology in the field of sound synthesis.
A range of potential future developments is also presented.

@@

% Essential background information
In this work it is assumed, without loss of generality, that one is interested in retrieving the mathematically ill defined \parencite{2013Mengempirical} temporal envelope of a arbitrary physical sound wave, of which only a recording is available.

Naturally, due to the necessary truncations both in time and in amplitude, as well as measure errors introduced by the instruments used to record the signal, it is not possible to characterise exactly the original, continuous wave, from its digital counterpart.

The best representation one can hope to get of the original wave is, then, the original recording. It is sensible, thus, to characterise the envelope of that wave in terms of a subset of the exact samples that comprise the digital representation.

To that end, this work will emphasize the concept of frontier, as the points of the digital wave that mark the wave's upper and lower boundaries; from those points an envelope, conforming to the vague definitions of uniqueness and smoothness, can be trivially constructed.

We start by defining a pulse as a series of consecutive samples in a discrete wave with the same sign. More formally, let $ W[i] \in \mathbb{R} \ \forall \ i \in \mathbb{N}_0, \ i < n \in \mathbb{N}_0 $ be a real, discrete and finite signal indexed, without loss of generality, over a adjacent subset of the natural numbers starting at zero. This definition relates closely to the concept of an array or vector in most programming languages, and is used in the interest of simplicity.

@@

In other words, each pulse is a unique subset of $ W $ with its members being the maximum sequence of samples in $ W $ with the same signal and $ P[0] \cup P[1] \cup \cdots \cup P[q-1] = W $. It is convenient to attribute to each pulse a pair of coordinates: those will be the coordinates of the sample with the maximum absolute amplitude in each pulse, being denoted by $ P[j]_x $ and $ P[j]_y $; additionally, the length of a pulse will be denoted by $ P[j]_l $.

Naturally, each pulse can be classified as positive or negative, based on the sign of its samples, which must be the same.
For a continuous function, then, a pulse would correspond to the pieces between its roots, and would be equal to half the cycle of a sinusoid. Thus, in our analogy, two consecutive pulses, naturally with opposing signs, have the potential do delimitate a pseudo-cycle, as illustrated in figure \ref{fig:Pulses}. For that to be true, however, the pulses need to make part of the frontiers of the signal, as will be defined later.

@@

To do so we will define the positive (upper) frontier and negative (lower) frontier as the set of points $ (i, |W[i]|) $ that are, necessarily but not sufficiently, the points of maximum absolute amplitude of their respective pulses. The set of all these points such that $ W[i] > 0 $ will be called the positive, piecewise linear frontier and will be denoted by $ F_+ $; similarly, $ F_- $ will denote the negative frontier of $ W $, defined by the subset of points $ (i, W[i]) $ where $ W[i] < 0 $, as is shown in figure \ref{fig:Frontiers}. To identify such points, we need to address the different units in which the amplitude and horizontal distance of our wave are expressed.

@@

To make abscissa and ordinate units consistent for the computation of the frontier, and proceed with a geometric treatment of the subject, we consider the set of positive and negative pulses separately. For each set, we multiply all pulses lengths by the average length of the pulses in the set, after having divided by the average of the pulses' amplitudes, putting both axes in the same unit $i$, related to time by the equation $i = t \ \mathit{fps}$.

Formally, supposing we have $ m $ pulses with the same sign, we scale each pulse's vertical coordinate by $s$, where:

\begin{equation} \label{eq:scale}
  s = \frac{\sum_{i=1}^{m} P[i]_l}{\sum_{i=1}^{m} P[i]_y}
\end{equation}

@@

Since an edge defines only one direction, there remains the need for a direction of reference. As we are interested in the relative curvature, it is natural to use the average direction of all positive or negative vectors, $ \vec{\overline{v}}^\pm = \frac{1}{m-1} \sum_{l=0}^{m-1} \vec{v}_l^\pm $, so as to measure the relative change of direction of each edge. This procedure, as well as the circles generated by it, are illustrated in figure \ref{fig:Curvatures}.

The angle $ \theta $ between the average direction and each of the vectors can be obtained via the relation between the slopes of the respective lines, as in the expression $ \theta = \arctan( (m - \overline{m}) / (1 + m \overline{m}) ) $. The radii of the equivalent circles are thus given by $ r_l = x_l / \sin (\theta_l) $ with the curvature given by $\kappa_l =1/r_l=\sin (\theta_l) / x_l$. After some algebra, one can write:

\begin{equation} \label{eq:radius}
  r_l = \left\lvert \frac{x_l \sqrt{\overline{m}^2 + 1} \ \sqrt{x_l^2 + y^2}}{\overline{m} x_l - y} \right\rvert
\end{equation}

@@

\section{Results}

Figure \ref{fig:FullFrontiers} illustrates the frontiers obtained to six diverse discrete sound waves, as well as a detail view of a shorter segment for each wave. All waves are records of physical sounds, chosen to represent the applicability of the algorithm in real world scenarios.

@@

\begin{figure}[ht!]
  \centering
    \includegraphics[width=0.8\linewidth]{07FullFrontiers.pdf}
  \caption{Positive and negative frontiers of 6 digital waves, as extracted by the algorithm here presented. For each wave, the region highlighted in black is shown in detail besides the whole wave. For each wave the horizontal axis is the sample number $i$, while the vertical axis is the normalized amplitude. All 6 waves were recorded at 44100 $ \mathit{fps} $.}
  \label{fig:FullFrontiers}
\end{figure}

@@

Figures \ref{fig:xRobustness} and \ref{fig:yRobustness} illustrate the computational resilience of the algorithm for scaling in both the abscissa and ordinate axes, respectively. This is due to the normalization scheme presented in equation \ref{eq:scale}.

Although important from a theoretical standpoint, this step has a special computational relevance in the case of digital sound waves, as they tend to be composed of a high number of samples in moderate amplitudes, generally normalized between -1 and 1, leading to a numerical disparity between axes.

\begin{figure}[ht!]
  \centering
    \includegraphics[width=0.8\linewidth]{10xRobustness.pdf}
  \caption{Minimum and maximum scales of the abscissa before computational stability issues arose, for a recording of a drumkit tom.  $ \text{B} = 10^6 $.}
  \label{fig:xRobustness}
\end{figure}

\begin{figure}[ht!]
  \centering
    \includegraphics[width=0.8\linewidth]{09yRobustness.pdf}
  \caption{Minimum and maximum values of the amplitude before computational stability issues arose, for a recording of a drumkit tom.}
  \label{fig:yRobustness}
\end{figure}

% Comparison with results in other works
Direct comparison with other algorithms is made difficult by the unavailability of digital implementations of such works, many designed to process analog signals \parencite[e.g.,][]{2018AssefModeling}.

\section{Discussion}

This work hopes to fill a gap identified by the authors in the context of sound synthesis, where the lack of procedures for the accurate identification of the temporal envelope of arbitrary waves consists of an obstacle to the complete description, and eventual manipulation of signals.

While relevant in its own accord, the procedure here presented isolates the individual cycles of a wave, pinpointing them in the time domain. Due to that, a rich set of theoretical advancements, of whose an example was given in the preceding section, can be built upon this initial building block.

With the knowledge of the exact position and characteristics of the cycles of a wave, one can investigate its evolution in time, in terms of frequency content for example, free from the potencial interference of a modulating wave; in other words, one can have a very satisfactory view of the instantaneous frequencies and phase of an arbitrary wave and its evolution in time, providing a way to bridge the gap between time domain and frequency domain interpretations.

\section{References}

\printbibliography[heading=none]
